Extending K-Means Clustering with Ptolemy’s Inequality

### 2024-10-02 content/4_contribution.tex
proofreading

\section{Contribution: Applying Ptolemy's Inequality to the Elkan's Algorithm}
\label{sec:contrib}

@@

\subsection{Ptolemaic Bounds}
Recall that the triangle inequality relates the distances between three points,
while Ptolemy's inequality does so for four points.
At the end of an iteration of Elkan's algorithm (Step~5~to~7 in \autoref{alg:elkan}),
the distance bounds between the data points $x_i$ and the new center positions are calculated $c'_j$.
Therefore, the three points used in the calculation of the bounds are the aforementioned two points and $c_j$, the cluster center calculated in the previous iteration.

@@

@@

\subsection{Integration}

>>>>

\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}
\label{sec:contrib}

@@

\subsection{Ptolemaic Bounds}
Recall that the triangle inequality relates the distances between three points,
while Ptolemy's inequality does so for four.
At the end of an iteration of Elkan's algorithm (Step~5~to~7 in Alg.~\autoref{alg:elkan}),
the distance bounds between the data points $x_i$ and the new center positions $c'_j$ are calculated.
Therefore, the three points used in the calculation of the bounds are the aforementioned two points and $c_j$, the cluster center calculated in the previous iteration.

@@

\subsection{Integration}
To integrate the novel bounds into the existing framework of Elkan's algorithm,
only Steps~6 and~7 in Alg.~\autoref{alg:elkan} need to be updated to use the new bounds.
It is possible to only replace either the upper or lower bound, which leads to a hybrid solution that is further explored in \autoref{sec: results}.

As the Ptolemaic bounds require bounds from previous iterations, the first iteration of the algorithm is always conducted with Elkan's original bounds.

@@

### 2024-10-02 main.tex
proofreading

%intro
Clustering is a fundamental data-analytical operation in the domain of data science, where the basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities, while objects in different clusters have significant differences. The collection of clusters reflects the inherent structure of the database and is denoted as clustering.

@@

The Elkan algorithm \cite{DBLP:conf/icml/Elkan03}, as one prominent representative of these algorithms particularly suited for high-dimensional scenarios, maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center. While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to make use of the Ptolemy inequality instead in order to increase the pruning performance. With this extension, we do not aim to ultimately compete with the state of the art in k-means clustering, but to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations. We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.

This short paper is structured as follows: Section \ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation. Section \ref{sec:contrib} then proposes how to apply Ptolemy's inequality to the aforementioned algorithm. The results of our preliminary performance evaluation are summarized in Section \ref{sec: results} while Section \ref{sec: conclusions} concludes this paper with a short outlook on future work.

@@

\todo{
	Experiments to show:
	- iris, wine, gaussian, uniform
	(moons, circles aren't a good fit for k-means -- is that reason enough to exclude them?)
}

@@

\pagebreak
%%% Angabe der .bib-Datei (ohne Endung) / State .bib file (im Falle der Nutzung von BibTeX)
\todo{re-enable citations}
\bibliography{references,bibliography_thesis,bibliography_iiwas}
%\printbibliography % im Falle der Nutzung von biblatex
\end{document}

>>>>

%intro
Clustering is a fundamental data-analytical operation in the domain of data science.
The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
meanwhile objects in different clusters should have significant differences.
The collection of clusters reflects the inherent structure of the database and is denoted as clustering.

@@

The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
With this extension, we do not aim to compete with the state of the art in k-means clustering;
rather we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations. We believe that our research findings offer promising directions for future research and new research impulses for data scientists and practitioners alike.

This paper is structured as follows:
Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of acceleration the algorithmic computation.
In Section~\ref{sec:contrib}, we proposes how to apply Ptolemy's inequality to the aforementioned algorithm as our main contribution.
The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.

@@

\todo{
	Experiments to show:
	- iris, wine, Gaussian, uniform
	(moons, circles aren't a good fit for k-means -- is that reason enough to exclude them?)
}

@@

%\bibliography{references,bibliography_thesis,bibliography_iiwas}
\end{document}
